---
title: "R_basics_Assignment_Radchenko"
author: "Anna Radchenko"
date: "January 14, 2020"
output:
  html_document: default
  pdf_document: default
---



```{r}
tgpp <- read.csv("~/R/R_basics/tgpp.csv") #Start of Lesson 1
```

#What are the names of the columns in this dataset?
```{r}
colnames(tgpp)
```

#How many rows and columns does this data file have?
```{r}
dim(tgpp)
```
#4080 Rows and 11 columns

#What kind of object is each data column? Hint: checkout the function sapply().
```{r}
sapply(tgpp, class) #don't have to put anything in the function, just name them!
```

#What are the values of the the datafile for rows 1, 5, and 8 at columns 3, 7, and 10
```{r}
tgpp[c(1,5,8),c(3,7,10)]
 #row then column
```

```{r}
SR <- plot(richness~scale, data = tgpp, xlab = 'Scale (m^2)', ylab = 'Richness', main = 'Scale vs. Richness', col = '#2E9FDF')
pdf('~/R/R_basics/Species_richness.fig1.pdf')
dev.off()
```



#What happens to your plot when you set the plot argument log equal to 'xy'. plot(..., log='xy')
```{r}
plot(richness~scale, data = tgpp, xlab = 'log (Scale (m^2))', ylab = 'log(Richness)', main = 'Scale vs. Richness', col = 2, log = 'xy')
```
#Changes both axis to the log scale
