---
title: "Spatial Models Assignment"
author: "Anna Radchenko"
date: "February 11, 2020"
output: html_document
---

Collaborator: Emily Parsons
```{r}
library(vegan)
data(BCI)
## UTM Coordinates (in metres)
BCI_xy = data.frame(x = rep(seq(625754, 626654, by=100), each=5), 
                    y = rep(seq(1011569,  1011969, by=100), len=50))
head(BCI)
```

```{r}
head(BCI_xy)
plot(BCI_xy)
```
```{r}
sr = apply(BCI, 1, function(x) sum(x > 0))
hist(sr) #calculating species richness for each site
```
```{r}
plot(BCI_xy, cex = sr/max(sr))
```

```{r}
col_brks = hist(sr, plot=F)$breaks
col_indices = as.numeric(cut(sr, col_brks))
cols = rev(terrain.colors(length(col_brks)))
plot(BCI_xy, cex=2, pch=19, col=cols[col_indices])
```
```{r}
sr_dist <- dist(sr)
xy_dist <- dist(BCI_xy)

max_dist = max(xy_dist) / 2

# plot result
plot(xy_dist, sr_dist)
abline(lm(sr_dist ~ xy_dist), lwd=3, col='red')
lines(lowess(xy_dist, sr_dist), lwd=3, col='pink')
abline(v = max_dist, col='red', lwd=3, lty=2)
```
Doesn't really look like the residuals are spatially dependent/autocorrelated. Slope of the line is almost 0 even in the lowess. 

```{r}
colSums(BCI)
```
Faramea.occidentalis 1717 chosen as the abundant/common species
Sapium.gladulosum 17 chosen as the rare species

```{r}
common <- BCI$Faramea.occidentalis
rare <- BCI$Sapium.glandulosum
common <- data.frame(common)
rare <- data.frame(rare)
```

```{r}
abundc = apply(common, 1, sum)
hist(abundc)
```


```{r}
col_brks = hist(abundc, plot=F)$breaks
col_indices = as.numeric(cut(abundc, col_brks))
cols = rev(terrain.colors(length(col_brks)))
plot(BCI_xy, cex=2, pch=19, col=cols[col_indices])
```


```{r}
abundc_dist <- dist(abundc)
xy_dist <- dist(BCI_xy)

max_dist = max(xy_dist) / 2

# plot result
plot(xy_dist, abundc_dist)
abline(lm(abundc_dist ~ xy_dist), lwd=3, col='red')
lines(lowess(xy_dist, abundc_dist), lwd=3, col='pink')
abline(v = max_dist, col='red', lwd=3, lty=2)
```
It does not look like there is much autocorrelation or spatial dependence for the common species Faramea occidentalis.The slope of the residuals is almost zero along an increasing spatial distance. 

```{r}
abundr = apply(rare, 1, sum)
hist(abundr)
```

```{r}
col_brks = hist(abundr, plot=F)$breaks
col_indices = as.numeric(cut(abundr, col_brks))
cols = rev(terrain.colors(length(col_brks)))
plot(BCI_xy, cex=2, pch=19, col=cols[col_indices])
```

```{r}
abundr_dist <- dist(abundr)
xy_dist <- dist(BCI_xy)

max_dist = max(xy_dist) / 2

# plot result
plot(xy_dist, abundr_dist)
abline(lm(abundr_dist ~ xy_dist), lwd=3, col='red')
lines(lowess(xy_dist, abundr_dist), lwd=3, col='pink')
abline(v = max_dist, col='red', lwd=3, lty=2)
```


```{r}
obs_cor_c <- cor(xy_dist, abundc_dist)
obs_cor_c
```

```{r}
# carry out the same analysis using the function mantel()
abundc_mantel = mantel(xy_dist, abundc_dist)
abundc_mantel
```

```{r}
# carry out a permutation test for significance:
nperm = 1000
null_cor = obs_cor_c
for (i in 2:nperm) {
    # shuffle the rows of the spatial coordinates
    tmp_xy = BCI_xy[sample(nrow(BCI_xy)), ]
    # correlation between the shuffled spatial coordinates and sr_dist
    null_cor[i] = cor(dist(tmp_xy), abundc_dist)
}
# compute the p-value
sum(null_cor >= obs_cor_c) / nperm 
```

```{r}
# compare the two approaches graphically using stacked boxplots
boxplot(list(null_cor, abundc_mantel$perm), horizontal = T, boxwex = 0.5,
        names = c('mine', 'theirs'), xlab='Correlation')
abline(v=obs_cor_c, col='red')
```
These figures and analyses indicate that there is no significant difference between a random spatial distribution and the observed distribution for Faramea occidentalis. Therefore, this common species is not spatially dependent. 

```{r}
obs_cor_r <- cor(xy_dist, abundr_dist)
obs_cor_r #how correlated is distance and the residuals, low number means not very correlated
```

```{r}
# carry out a permutation test for significance:
nperm = 1000
null_cor = obs_cor_r
for (i in 2:nperm) {
    # shuffle the rows of the spatial coordinates
    tmp_xy = BCI_xy[sample(nrow(BCI_xy)), ]
    # correlation between the shuffled spatial coordinates and sr_dist
    null_cor[i] = cor(dist(tmp_xy), abundr_dist)
}
# compute the p-value
sum(null_cor >= obs_cor_r) / nperm
```

```{r}
# carry out the same analysis using the function mantel()
abundr_mantel = mantel(xy_dist, abundr_dist)
abundr_mantel
```

```{r}
# compare the two approaches graphically using stacked boxplots
boxplot(list(null_cor, abundr_mantel$perm), horizontal = T, boxwex = 0.5,
        names = c('mine', 'theirs'), xlab='Correlation')
abline(v=obs_cor_r, col='red')
```
These figures and analyses indicate that there is no significant difference between a random spatial distribution and the observed distribution for Sapium.gladulosum. Therefore, this common species is not spatially dependent. 
```{r}
sp_ids = c("Cordia.lasiocalyx", "Hirtella.triandra",
           "Picramnia.latifolia", "Quassia.amara",
           "Tabernaemontana.arborea", "Trattinnickia.aspera", 
           "Xylopia.macrantha")
sp_ids
```
```{r}
sp_a <- BCI$Drypetes.standleyi
sp_b <- BCI$Cordia.lasiocalyx
sp_c <- BCI$Hirtella.triandra
sp_d <- BCI$Picramnia.latifolia
sp_d <- BCI$Quassia.amara
sp_e <- BCI$Tabernaemontana.arborea
sp_f <- BCI$Trattinnickia.aspera
sp_g <- BCI$Xylopia.macrantha

sp_ids <- cbind(sp_a,sp_b,sp_c,sp_d,sp_e,sp_f, sp_g)
sp_ids <- data.frame(sp_ids)
  
```




```{r}
sp <- apply(sp_ids, 1, sum)
sp
```

```{r}
sp_dist <- dist(sp)
xy_dist <- dist(sp_ids)

max_dist = max(xy_dist) / 2

# plot result
plot(xy_dist, sp_dist)
abline(lm(sp_dist ~ xy_dist), lwd=3, col='red')
lines(lowess(xy_dist, sp_dist), lwd=3, col='pink')
abline(v = max_dist, col='red', lwd=3, lty=2)
```

```{r}
sp_mantel = mantel(xy_dist, sp_dist)
sp_mantel
```

```{r}
## compute bray curtis distance for the community matrix
comm_dist = vegdist(sp_ids)
plot(xy_dist, comm_dist)
abline(lm(comm_dist ~ xy_dist), lwd=3, col='red')
lines(lowess(xy_dist, comm_dist), lwd=3, col='pink')
lines(lowess(xy_dist, comm_dist, f=0.1), lwd=3, col='blue')

abline(v = max_dist, col='red', lwd=3, lty=2)
```
Residuals show spatial dependence.

```{r}
sp_corlog = mantel.correlog(sp_dist, xy_dist)
comm_corlog = mantel.correlog(comm_dist, xy_dist)
sp_corlog
```

```{r}
comm_corlog
```

```{r}
par(mfrow=c(1,2))
plot(sp_corlog)
mtext(side=3, 'Species Abundance')
abline(v = max_dist, col='red', lwd=3, lty=2)
plot(comm_corlog)
mtext(side=3, 'Community Composition')
abline(v = max_dist, col='red', lwd=3, lty=2)
```

```{r}
sp_dat = data.frame(sp_ids, BCI_xy)
head(sp_dat)

sp_lm = gls(sp_a ~ sp_b, data = list(sp_dat))
sp_lm
summary(sp_lm)
plot(Variogram(sp_lm, form= ~ x + y)
``` 

```{r}
res = residuals(sp_lm)
plot(dist(sp_dat[, c('x', 'y')]), dist(res))
lines(lowess(dist(sp_dat[, c('x', 'y')]), dist(res)), col='red', lwd=2)
abline(v = max_dist, col='red', lwd=3, lty=2)
```

```{r}
sp1_lm = gls(sp_a ~ sp_b + sp_c + sp_d + sp_e + sp_f + sp_g, data = list(sp_dat))
sp1_lm
summary(sp1_lm)
plot(Variogram(sp1_lm, form= ~ x + y)
```

```{r}
res = residuals(sp1_lm)
plot(dist(sp_dat[, c('x', 'y')]), dist(res))
lines(lowess(dist(sp_dat[, c('x', 'y')]), dist(res)), col='red', lwd=2)
abline(v = max_dist, col='red', lwd=3, lty=2)
```

```{r}
sp_exp = update(sp1_lm, corr=corExp())
# examine fit of error model to the raw model residuals
# note this function defaults to displaying pearson standardized residuals
# resType='p' or resType='pearson'
plot(Variogram(sp_exp, maxDist = max_dist))

```


```{r}
plot(Variogram(sp_exp, resType='normalized', maxDist = max_dist))
```

```{r}
sp_exp_nug = update(sp_exp, corr=corExp(, nugget=T))
plot(Variogram(sp_exp_nug, maxDist = max_dist))
```

```{r}
plot(Variogram(sp_exp_nug, resType='n', maxDist = max_dist))
```

```{r}
sp_rat_nug = update(sp_lm, corr=corRatio(, nugget=T))
# examine fit of error model to model residuals
plot(Variogram(sp_rat_nug, maxDist = max_dist))
```

```{r}
plot(Variogram(sp_rat_nug, resType='n', maxDist = max_dist))
```

```{r}
anova(sp_lm, sp_exp, sp_exp_nug, sp_rat_nug, test=F)
```
Including the spatial error term made the models better. According to the anova, the best model was the spatial experimental model. The model including the error is better because the model helps to explain the spatial portion of the error.



